Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-15 numerical-starformation.tex
intermediate checkin.

While the physics discussed in other sections of this paper is crucial to the study of
cosmological structure formation, most cosmological observations in
the infrared, optical, and ultraviolet wavelengths are of
stellar populations.  Furthermore, stars eject energy and metal-
enriched gas throughout their lives, drastically modifying their own 
environments' cooling properties.  Broadly speaking, the formation of galaxies and clusters cannot be completely
modeled without including the feedback of energy and metals.  In particular,
it is thought that feedback is crucial for suppressing the large numbers of
dwarf-like galaxies that CDM theories predict 
\citep{1991ApJ...381...14L,1991ApJ...379...52W}.  An early
burst of star formation could remove a large fraction of cold gas from such systems
\citep{1978MNRAS.183..341W,1991ApJ...367...45C}.  Also, the unexpectedly 
low luminosities of small clusters and
groups (relative to the richest clusters) is commonly explained through feedback
\citep{1991ApJ...383..104K}.  Energy ejected during the formation of the cluster ellipticals 
lowers the central density and hence the X-ray luminosity of such clusters 
 \citep{1997ApJ...484L..21C}.  For all of these reasons, it is
 critical to include star formation and the feedback of energy and
 metals into the interstellar and intergalactic media in cosmological simulations.

We have extended the \enzo\ code to include multiple models for star formation
and feedback.  It is difficult to directly simulate the formation of individual
stars in the context of simulations of galaxy formation and evolution due
to the immense computational cost.  Therefore, we adopt a parametric method which attempts
to model star forming regions in a plausible but heuristic way.  
%One model is based on Kravtsov's method  \citep{2003ApJ...590L...1K}, and the other is 
Our model is based on the method of  \citet{CO1992}, which 
has been modified for use in an AMR code.  The basic
ideas behind the methods are straightforward and observationally motivated.
Similar (though somewhat more advanced) methods have been employed for the 
smoothed particle hydrodynamics
(SPH) method, most recently by Springel \& Hernquist 
\citep{sh03a,sh03b,hs03}. This 
method assumes a multiphase IGM and has been shown to accurately reproduce the cosmic star
formation rate, and will be implemented into \enzo\ in the near future.

%In the following sections we will detail both the Kravtsov and Cen \& Ostriker star formation methods separately.  
Tests of the star formation and
feedback algorithms will not be shown, since this work is currently in
progress (and an extension of the AMR/SPH code comparison described in
\citet{2005ApJS..160....1O}).
%
%\subsubsection{The Kravtsov star formation and feedback algorithm}

%\subsubsection{The Cen \& Ostriker star formation algorithm}

The Cen \& Ostriker method is a heuristic model of star formation
on galactic scales.  This method, first described by
\citet{CO1992}, is similar in some ways to the Kravtsov algorithm
but has more complex criteria for determining where stars should
be formed.  In this method, cells that form stars must have a 
baryon overdensity higher than some threshold 
$\rho_b/\bar{\rho}_b \geq \eta$ 
where $\rho_b$ is the baryon density in that cell, 
$\bar{\rho_b}$ is the mean baryon density in the simulation,
and $\eta$ is the user-defined overdensity threshold.
Additionally, the gas in the cells must be contracting, 
cooling rapidly, and gravitationally unstable, e.g.:

@@

>>>>

\subsubsection{Overview}
% Overview of methodology

While the physics discussed in other sections of this paper is crucial
to the study of cosmological structure formation, most observations of
galaxies in the infrared, optical, and ultraviolet wavelengths are of
stellar populations.  Furthermore, stars eject energy and metal-
enriched gas throughout their lives, drastically modifying their own
environments' cooling properties.  Broadly speaking, the formation of
galaxies and galaxy clusters cannot be completely modeled without
including the feedback of energy and metals from stellar populations.
In particular, it is thought that feedback is crucial for suppressing
the large numbers of dwarf-like galaxies that CDM theories predict
\citep{1991ApJ...381...14L,1991ApJ...379...52W}.  An early burst of
star formation could remove a large fraction of cold gas from such
systems \citep{1978MNRAS.183..341W,1991ApJ...367...45C}.  Also, the
unexpectedly low luminosities of small clusters and groups (relative
to the richest clusters) is commonly explained through feedback
\citep{1991ApJ...383..104K}.  Energy ejected during the formation of
the cluster ellipticals lowers the central density and hence the X-ray
luminosity of such clusters \citep{1997ApJ...484L..21C}.  For all of
these reasons, it is critical to include star formation and the
feedback of energy and metals into the interstellar and intergalactic
media in cosmological simulations.

Modeling star formation in the context of galaxy formation forces one
to confront two fundamental problems. The dynamical range required to
study galaxy formation and evolution is substantial, and attempting to
directly resolve the formation of individual stars in the context of a
Milky Way-type galaxy is computationally unfeasible.  Also, a strong
theoretical understanding of the details of star formation is
currently lacking.  Due to these challenges, models of star formation
in cosmological simulations are largely phenomenological, and the
``stars'' that are formed are in most cases meant to represent an
ensemble stellar population (although not all; see
Section~\ref{sec:starform_pop3}).
% BWO: I use 'phenomenological' right above this, but could probably
% go for 'parametric' or 'heuristic' if people object to my usage.

The \enzo\ code includes approximately ten models for star formation
and feedback.  Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with the
same mass is placed in the center of that cell, with the same momentum
vector as the removed gas.  This star particle is then allowed to
inject mass, momentum, thermal energy, metals, and possibly magnetic
fields and/or cosmic ray populations into its local environment.  In
general, the particle is treated as an ensemble of stars, with
feedback properties occuring over time according to the assumed
initial mass function of the stellar population.

In the following sections, we describe some of the more widely-used star formation
and feedback methods used in \enzo.  We note that similar methods have
been employed in many other codes used for galaxy formation, with
comparable implementations for both star formation and feedback in
other grid-based codes.  With regards to particle-based codes, star
formation is broadly similar in implementation, though feedback is
typically implemented in a very different way due to the Lagrangian
nature of the method \citep[see, e.g.,][]{sh03a,sh03b,hs03}.

% Specific example: Cen & Ostriker
\subsubsection{Cen \& Ostriker}
\label{sec:starform_cen}

The Cen \& Ostriker method is a heuristic model of star formation
on galactic scales.  This method, first described by
\citet{CO1992}, assumes that star particles

s similar in some ways to the Kravtsov algorithm
but has more complex criteria for determining where stars should
be formed.  In this method, cells that form stars must have a 
baryon overdensity higher than some threshold 
$\rho_b/\bar{\rho}_b \geq \eta$ 
where $\rho_b$ is the baryon density in that cell, 
$\bar{\rho_b}$ is the mean baryon density in the simulation,
and $\eta$ is the user-defined overdensity threshold.
Additionally, the gas in the cells must be contracting, 
cooling rapidly, and gravitationally unstable, e.g.:

@@

% Specific example: Kravtsov
\subsubsection{Kravtsov}
\label{sec:starform_kravtsov}

% Specific example: H2-regulated
\subsubsection{H2-regulated}
\label{sec:starform_H2reg}

% Specific example: Pop III
\subsubsection{Population III}
\label{sec:starform_pop3}

%\subsubsection{The Kravtsov star formation and feedback algorithm}
